---
title: "ERCOT Price Gouging"
author: "Megha Joshi"
date: 2021-02-23
categories: ["R"]
tags: ["ERCOT", "ggplot", "Texas freeze", "power outage"]
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 
During the crisis brought on by government failure and capitalism last week in Texas that cause millions of people to suffer in freezing temperatures and suffer without water, some Texans [received egregious electricity bills](https://www.washingtonpost.com/nation/2021/02/21/texas-high-electric-bills/). 

I found the historical data on Electric Reliability Council of Texas (ERCOT) settlement point prices [here](http://mis.ercot.com/misapp/GetReports.do?reportTypeId=13060&reportTitle=Historical%20DAM%20Load%20Zone%20and%20Hub%20Prices&showHTMLView=&mimicKey). ERCOT manages the power grid system in Texas. According to [Quick Facts](http://www.ercot.com/content/wcm/lists/144926/ERCOT_Quick_Facts_8818.pdf) from ERCOT, "it also performs financial settlement for the competitive wholesale bulk-power market and administers retail switching for seven million premises in competitive choice areas." 

Below I plot Daily-Ahead Market Settlement Point Prices (DAM SPP). According to [ERCOT website](http://www.ercot.com/mktinfo/dam), "The Day-Ahead Market (DAM) is a voluntary, financially-binding forward energy market. The DAM matches willing buyers and sellers, subject to network security and other constraints..." The data from ERCOT contains DAM SPPs by hour of the day. I plot average DAM SPP per day across hours. The DAM SPPs increased to nearly $9,000 during the crisis last week.


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(scales)

ercot_dat <- read_csv("https://raw.githubusercontent.com/meghapsimatrix/datasets/master/ercot/ercot_dam_spp.csv")

glimpse(ercot_dat)

# not collapsing across hours
summary(ercot_dat$settlement_point_price)
```



```{r}
ercot_dat %>%
  mutate(delivery_date = mdy(delivery_date)) %>%
  group_by(delivery_date) %>%
  summarize(mean_spp = mean(settlement_point_price)) %>%
  ggplot(aes(x = delivery_date, y = mean_spp)) + 
  geom_point(alpha = .5) + 
  geom_line(color = "blue") + 
  scale_y_continuous(breaks = seq(0, 10000, 1000), 
                     label = comma) +
  scale_x_date(breaks = pretty_breaks(10)) +
  labs(x = "Date", y = "Average Day-Ahead Market Settlement Point Price") + 
  theme_bw()
```

Make utilities public!