---
title: "Propensity Scores Analysis with Multiply Imputed Data"
author: "Megha Joshi"
date: 2020-06-23
categories: ["R"]
tags: ["propensity scores", "missing data", "multiple imputation" , "causal"]
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Work in progress...

In this post, I walk through steps of running propensity scores when there is missingness in the covariate data. Particularly, I look at multiple imputation and ways to condition on propensity scores estimated on imputed data. The code builds on my earlier [post](https://meghapsimatrix.com/post/missing_dat/) that goes over different ways to handle missing data when conducting propensity score analysis.

When using multiple imputations, Hill (2004) and Mitra and Reiter (2016) examined two distinct ways to condition on the propensity scores:

## Multiple Imputation Across (MI Across)

This approach involves creating _m_ imputed datasets and then estimating propensity scores within each of the datasets and then averaging each unit's _m_ propensity scores across the _m_ datasets [@hill_2004]. Stratification, matching or IPW can be implemented using these averaged propensity scores [@hill_2004]. Outcome models that include covariates will need to use the weights or strata derived from the averaged propensity scores and the _m_ sets of covariate values. The weighted regression estimates will then need to be pooled. 

## Multiple Imputation Within (MI Within)

This approach involves creating _m_ imputed datasets and then estimating propensity scores within each of the datasets [@hill_2004]. Instead of averaging the propensity scores across the datasets, this method entails conditioning on the propensity scores within the datasets and running the outcome analyses within each dataset [@hill_2004]. The separate regression estimates have to be pooled. 


# Read in the data

Below, I show how to implement the Across and Within methods to estimate the ATT. The data that I use here is from High School and Beyond (HSB) longitudinal study used by Rosenbaum (1986) to analyze the effect of dropping out of high school on later math achievement. 

```{r, warning = F, message = F}
library(tidyverse)
library(mice)
library(twang)
library(estimatr)
library(broom)
library(naniar)
library(knitr)
```

```{r, warning = F, message = F}
hsls_dat <- read_csv("https://raw.githubusercontent.com/meghapsimatrix/datasets/master/causal/HSLS09_incomplete.csv") %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(vars(repeated_grade, IEP), as.factor) %>%
  select(-working_T3) # an outcome we don't care about
```


Below, using `gg_miss_var` from the `naniar` package, I visualize the percent of data that are missing for each of the variables. The treatment variable, `drop_status` has no missing data. The outcome variable, `math_score_T2`, however, does have around 10% missing data. 

```{r}
gg_miss_var(hsls_dat, show_pct = TRUE)
```


# Multiple imputation using `mice`

Here, I impute the data using mice and set the number of imputations and itereations to 10 each and also set the seed. Then, I save the result as an `RData` file. For imputation method, I let mice run the default methods: predicitive mean matching for continuous variables, Bayesian logistic regression for binary variables and Bayesian polytomous regression for multinomial variables. For more information on methods for imputation, please see [Chapter 6.2](https://stefvanbuuren.name/fimd/sec-modelform.html) from [Flexible Modeling of Missing Data](https://stefvanbuuren.name/fimd/), van Buuren (2018).

```{r, eval = F}
system.time(temp_data <- mice(hsls_dat, m = 10, maxit = 10, seed = 20200516))

save(temp_data, file = "temp_data.RData")
```


# Long format and propensity score model

I then load the saved RData file and then extract a long format data that contains each of the 10 imputed data stacked. 

```{r}
load("temp_data.RData")
imp_dat <- complete(temp_data, action = "long")
```


# Function to estimate propensity scores

Below I create a dataset with only the covariates and use the `paste()` function to create propensity score model equation. For the sake of this example, I only focus on including main effects of the covariates in the propensity score model. 

```{r}
covs <- imp_dat %>%
  select(sex:climate, math_score_T1)

equation_ps <- paste("drop_status ~ ", paste(names(covs), collapse = " + "))
equation_ps
```

Here, I create a function `estimate_ps()` that takes in an equation and a dataset, then runs logistic regression using `glm()`. Then the function adds logit of propensity scores and propensity scores as columns in the dataset. 

I then group the `imp_dat` by the imputation number and then run the `estimate_ps()` function on each of the imputed dataset using the `do()` function from `dplyr`.

```{r}
estimate_ps <- function(equation, dat){
  
  ps_model <- glm(as.formula(equation), family = binomial, data = dat)
  
  dat <- dat %>%
    mutate(ps_logit = predict(ps_model, type = "link"),
           ps = predict(ps_model, type = "response"))
  
  return(dat)
  
}


imp_dat_ps <- imp_dat %>%
  group_by(.imp) %>%
  do(estimate_ps(equation_ps, .)) %>%
  ungroup()
```


# Estimating weights using the across and within methods

Below I estimate ATT weights using the across and within methods. For the Across method, I use the average of the propensity scores across the imputed datasets to calculate weights. For the within method, I use the the propensity scores estimated within each imputed dataset to calculate weights.

The code below groups the imputed data by `.id` which is an identifier denoting each case. For each case, I summarize the mean of the propensity scores across the 10 imputed dataset. I then estimate the ATT weights using the averaged propensity scores for the Across method and the original propensity scores for the Within method. 


```{r}
imp_dat_ps <- imp_dat_ps %>%
  group_by(.id) %>%
  mutate(ps_across = mean(ps)) %>%
  ungroup() %>%
  mutate(att_wt_across = drop_status + (1 - drop_status) * ps_across/(1 - ps_across),
         att_wt_within = drop_status + (1 - drop_status) * ps/(1 - ps))


imp_dat_ps %>%
  select(.imp, ps, ps_across, att_wt_across, att_wt_within)
```


# Estimate ATT for across and within methods

Below I create a function to estimate the ATT. The function takes in an equation, a dataset, and weights as arguments. Then it runs a model using `lm_robust()` from the `estimatr` package. The standard errors of the regression coefficients are estimated using `HC2` type sandwich errors. I then clean up the results using `tidy()` from `broom`. 

```{r}
estimate_ATT <- function(equation, dat, wts){
  
  wts <- dat %>% pull({{wts}})
  
  model <- lm_robust(as.formula(equation), data = dat, weights = wts)
  
  res <- model %>%
    tidy() %>%
    filter(term == "drop_status") %>%
    select(term, estimate, se = std.error, dci_low = conf.low, ci_high = conf.high, df = df)
  
  return(res)
  
}
```

I set up an equation regressing the outcome variable, `math_score_T2` on drop status and on the main effects of all the covariates. I then run the `estimate_ATT()` function on each imputed data using `group_by()` and `do()`. The weights are different for the Across and Within methods. 

```{r}
equation_ancova <- paste("math_score_T2 ~ drop_status + ", paste(names(covs), collapse = " + "))
equation_ancova

across_res <- imp_dat_ps %>%
  group_by(.imp) %>%
  do(estimate_ATT(equation = equation_ancova, dat = ., wts = att_wt_across)) %>%
  ungroup()

within_res <- imp_dat_ps %>%
  group_by(.imp) %>%
  do(estimate_ATT(equation = equation_ancova, dat = ., wts = att_wt_within)) %>%
  ungroup()
```


# Pool the results 

Here, I create a function called `calc_pooled()` using formula by (1999) to pool the results across the imputations. The `mice` package has the `pool()` function to do the same thing but we would need to convert the imputed data back to `mids` object type and I just wanted to skip all that :D 

```{r}
calc_pooled <- function(dat, est, se, df){
  
  dat <- dat %>%
    mutate(est = dat %>% pull({{est}}),
           se = dat %>%pull({{se}}),
           df = dat %>% pull({{df}}))
  
  pooled <- dat %>%
    summarize(m = n(),
              B = var(est),  # between imputation var
              beta_bar = mean(est), # mean of estimated reg coeffs
              V_bar = mean(se^2), # mean of var - hc corrected   within imp var
              eta_bar = mean(df)) %>%   # mean of df
    mutate(
      
      V_total = V_bar + B * (m + 1) / m,  #between and within var est
      gamma = ((m + 1) / m) * B / V_total,  
      df_m = (m - 1) / gamma^2,
      df_obs = eta_bar * (eta_bar + 1) * (1 - gamma) / (eta_bar + 3),
      df = 1 / (1 / df_m + 1 / df_obs),
      
      # output
      se = sqrt(V_total),
      ci_lower = beta_bar - se * qt(0.975, df = df),
      ci_upper = beta_bar + se * qt(0.975, df = df)) %>%
    
    select(est = beta_bar, se, df, ci_lower, ci_upper) 
  
  return(pooled)
  
}

```

# Pooling

Below I use the `calc_pooled()` function to pool the results for each of the methods. 

## Across

```{r}
across_pooled <- calc_pooled(dat = across_res, est = estimate, se = se, df = df)
across_pooled %>%
  kable()
```


## Within

```{r}
within_pooled <- calc_pooled(dat = within_res, est = estimate, se = se, df = df)
within_pooled %>%
  kable()
```


# Which to use



# References
