---
title: "Nepal Earthquake"
author: "Megha Joshi"
date: 2019-11-20
categories: ["R"]
tags: ["Nepal", "Nepal Earthquake 2015", "leaflet", "gganimate", "ggplot"]
---

I wanted to analyze the data from the [April 2015 Nepal earthquake](https://en.wikipedia.org/wiki/April_2015_Nepal_earthquake) that resulted in around 10,000 deaths. I am using a [dataset](https://data.world/opennepal/1b7b5d6e-3c98-49f4-a884-a167c4040d3a) that I found in `data.world`. The data contains date, time, location and magnitude of the earthquake and the many aftershocks that followed. The data is updated as of June 2, 2015.


# Libraries

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(geojsonio)
library(broom)
library(gganimate)
library(leaflet)
```



# Read in the data

```{r, message=FALSE, warning=FALSE}
earthquake_dat <- read_csv("https://raw.githubusercontent.com/meghapsimatrix/Data_Visualization/master/data/earthquake-0-csv-1.csv") %>%
  mutate(lab = paste0(Epicentre, "; ", "Magnitude(ML): ", `Magnitude(ML)`),
         is_gorkha = if_else(`Magnitude(ML)` == max(`Magnitude(ML)`), "Gorkha", "Aftershocks"))

head(earthquake_dat)

# there is one entry where I think the lat and long are switched
summary(earthquake_dat$Latitude)
summary(earthquake_dat$Longitude)

# Gorkha seems like the lat and long are switched
(outlier <- earthquake_dat %>%
  filter(Latitude == max(Latitude)))

earthquake_dat <- earthquake_dat %>%
  mutate(Latitude = if_else(lab == outlier$lab & Date == outlier$Date, outlier$Longitude, Latitude),
         Longitude = if_else(lab == outlier$lab & Date == outlier$Date, outlier$Latitude, Longitude))

# Sindhupalchowk seems like the Longitude is wrong
earthquake_dat %>%
  filter(Longitude == min(Longitude))

sindhupalchowk <- earthquake_dat %>%
  filter(str_detect(Epicentre, "Sindhu"))

# Mean imputing based on other values for Sindhupalchowk
earthquake_dat <- earthquake_dat %>%
  mutate(Longitude = if_else(Longitude == min(Longitude), mean(sindhupalchowk$Longitude), Longitude))
```



# Leaflet 

Created using the `leaflet` package:

```{r}
leaflet(earthquake_dat) %>% 
  addTiles()  %>% 
  setView(lat = 27, lng = 85, zoom = 7) %>%
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addCircleMarkers(~Longitude, ~Latitude,
                   radius = ~`Magnitude(ML)`, popup = ~lab, stroke = FALSE, opacity = 2)
```


# Make a Map of Nepal

I got the code for the base map from [here](https://stackoverflow.com/questions/50859765/chloropleth-map-with-geojson-and-ggplot2).

```{r warning = FALSE, message = FALSE}
np <- geojson_read("https://raw.githubusercontent.com/mesaugat/geoJSON-Nepal/master/nepal-districts.geojson",  what = "sp")
np_dat <- tidy(np)


# plot
np_plot <- ggplot() +
  geom_polygon(data = np_dat, aes( x = long, y = lat, group = group)) 

np_plot
```

# Mapping on the Earthquake and Aftershocks

Now plotting the latitude and longitudes. Size indicates the magnitude of the earthquake.

```{r}
(np_earthquake <- np_plot + 
  geom_point(data = earthquake_dat, 
             aes(x = Longitude, y = Latitude, 
                 size = `Magnitude(ML)`), 
             color = "red", alpha = .5) + 
  labs(color = "") + 
  theme_void() +
  theme(legend.position = "none"))

```


# Animating

```{r}
np_animate <- np_earthquake +
  transition_states(Date) +
  labs(title = 'Date: {closest_state}') +
  enter_appear() +
  exit_disappear()

animate(np_animate)
```






