---
title: "Tidy Tuesday Horror"
author: "Megha Joshi"
date: 2019-10-21
categories: ["R"]
tags: ["tidy tuesday", "ggplot", "tidyverse"]
---


# Load the Data 

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(kableExtra)


# there were complete duplicated rows
dat <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-22/horror_movies.csv") %>%
  distinct(.) # removes complete dups

# check duplicates
dup_title <- dat %>%
  filter(duplicated(title) | duplicated(title, fromLast = TRUE)) %>%
  arrange(title)

# examined they seem different movies even though same title
dup_title %>%
  filter(duplicated(plot))

dup_title %>%
  filter(duplicated(release_date)| duplicated(release_date, fromLast = TRUE))

# The Jokesters seems to be a duplicate but with different rating and run time
# Deleting it for now
dat <- dat %>%
  filter(title != "The Jokesters (2015)")

```

# Genres

The genre column looked extremely messy so some `stringr` fun. Each film can be categorized into multiple genres. 

```{r}
dat <- dat %>%
  mutate(genres = str_remove_all(genres, "\\|"))  # remove |

dat_long <- dat %>% 
  separate_rows(genres, sep = " ")  # long format

# Just to check - looks okay - just 1 movie with no genre
table(dat_long$genres, useNA = "ifany")

dat_long <- dat_long %>%
  mutate(genres = fct_infreq(fct_lump(genres, n = 8))) # Factor keeping 8 most frequent categories and lumping the rest to Other and order the factor by frequency
```

## Table: Number of Films per Genre

```{r}
genre_count <- dat_long %>%
  filter(!is.na(genres)) %>%
  group_by(genres) %>%
  summarize(n = n()) %>%
  ungroup() 

kable(genre_count, format = "html", table.attr = "style = \"color: white;\"") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

## Bar Graph: Distribution of Genres

```{r, fig.height = 4, fig.length = 5}
genre_count %>%
  ggplot(aes(x = genres, y = n, fill = genres)) +
  geom_bar(stat = "identity") + 
  scale_y_continuous(labels = scales::comma) +  # y axis to have commas 
  scale_fill_brewer(palette ="BuPu", direction = -1) + # reverse order the palette
  theme_light() + 
  labs(x = "", y = "Count") + 
  theme(legend.position = "none")
```


# Review Rating by Release Year

Some of the years are `dmy` format, some just have the years. I am extracting the year and filling in any that didn't parse with the year value from the original release_date column. No missing values for year :) 

```{r, warning = FALSE, message = FALSE}
date_dat <- dat %>%
  mutate(date = dmy(release_date),
         yr = year(date),
         yr = ifelse(is.na(yr), release_date, yr))

table(is.na(date_dat$yr))

date_dat %>%
  select(release_date, date, yr) %>%
  filter(is.na(date)) %>%
  head()

  
date_dat %>%  
  ggplot(aes(x = yr, y = review_rating, fill = yr)) +
  geom_boxplot(alpha = .5) + 
  labs(x = "Release Year", y = "Review Rating") + 
  theme_light()  + 
  theme(legend.position = "none")
```


Looks like there is a slight increase in ratings for newer films.
